---
title: "FS_Linked_Unrelated"
output: html_document
date: "2025-06-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Packages

```{r}
library(dplyr)
library(readr)
```

# The 50,000 .ped files from the FS_Linked simulation are manipulated to simulate unrelated individuals.
# This is done by:
# 1) Breaking familial relationships by setting columns 3 (Mother ID) and 4 (Father ID) to 0.
# 2) Removing extra rows/columns added during simulation.
# 3) Standardising each .ped file to contain only two unrelated individuals:
#    - First 5 columns fixed
#    - Remaining columns: genotypes from each simulated .ped file


```{r}
# Set the directories, input directory is where the FS_Linked simulated .peds are stored and the output directory is where the manipulated .peds will be stored
in_dir  <- "/Users/huncho/Desktop/Forensic_Kinship_Project/FS_Simulation/FS_Linked"
out_dir <- "/Users/huncho/Desktop/Forensic_Kinship_Project/FS_Simulation/FS_Linked_Unrelated"

# Create the output directory if it doesn’t exist
if (!dir.exists(out_dir)) dir.create(out_dir)

# Grab all FS_Linked .ped files matching the pattern:
# "merlin-00003898-xxxxx-replicate.ped" (where xxxxx ranges from 00001 to 50000)
all_fns <- list.files(
  path       = in_dir,
  pattern    = "^merlin-00003898-\\d{5}-replicate\\.ped$",
  full.names = TRUE
)

# 2) Define the transformation as a function
process_ped <- function(in_ped_path, out_ped_path) {
  df <- read_delim(in_ped_path, delim = "\t", col_names = FALSE)
  
  df_mod <- df %>%
    
    # Set X3 and X4 (mother and father IDs) to 0 to break any pedigree structure
    mutate(
      X3 = 0,
      X4 = 0
    ) %>%
    
    # Shift genotypes from rows 3 and 4 to become the first two rows (representing two unrelated individuals)
    { tmp <- . 
      cols_to_shift <- 6:ncol(tmp)
      tmp[1, cols_to_shift] <- tmp[3, cols_to_shift]
      tmp[2, cols_to_shift] <- tmp[4, cols_to_shift]
      tmp
    } %>%
    
    # Drop all original rows except the newly reassigned unrelated pair (keep only rows 1 and 2)
    slice(-c(3,4,5)) %>%
    
    # Drop column X100 (presumably an extra column from simulation output)
    select(-X100)
  
  # Write the modified .ped file to the output directory
  write_delim(
    df_mod,
    out_ped_path,
    delim     = "\t",
    col_names = FALSE
  )
}

# Loop over all input .ped files and apply the transformation function
for (in_file in all_fns) {
  # construct the output filename by replacing ".ped" with "_UN.ped" (unrelated) 
  # e.g., merlin-00003898-00001-replicate.ped --> merlin-00003898-00001-replicate_UN.ped 
  base    <- basename(in_file) 
  base_no_ext <- tools::file_path_sans_ext(base)
  new_base <- paste(base_no_ext, "_UN.ped", sep = "")
  out_file <- file.path(out_dir, new_base)
  
  process_ped(in_file, out_file)
}
```

# Verify the structure of the 1st .ped file produced from the simulation and the manipulated .ped to verify correct manipulation
# Will also pick a random .ped out of the 50,000 to do the same

# Verify 1st simulated .ped File

```{r}
# Define file paths
Simulated_FS_Ped1_path <- "/Users/huncho/Desktop/Forensic_Kinship/94_SNP_Simulation/FS_Simulation/FS_Linked/merlin-00003898-00001-replicate.ped"

# Load the Linked file
Simulated_FS_Ped1 <- read_delim(Simulated_FS_Ped1_path, delim = "\t", col_names = FALSE)

# Display structure of the files
View(Simulated_FS_Ped1)
```

# Verify 1st manipulated .ped File

```{r}
# Define file paths
Manipulated_FS_Ped1_path <- "/Users/huncho/Desktop/Forensic_Kinship/94_SNP_Simulation/FS_Simulation/FS_Linked_Unrelated/merlin-00003898-00001-replicate_UN.ped"

# Load the Linked file
Manipulated_FS_Ped1 <- read_delim(Manipulated_FS_Ped1_path, delim = "\t", col_names = FALSE)

# Display structure of the files
View(Manipulated_FS_Ped1)
```

# Verify 27674 simulated .ped File

```{r}
# Define file paths
Simulated_FS_Ped27674_path <- "/Users/huncho/Desktop/Forensic_Kinship/94_SNP_Simulation/FS_Simulation/FS_Linked/merlin-00003898-27674-replicate.ped"

# Load the Linked file
Simulated_FS_Ped27674 <- read_delim(Simulated_FS_Ped27674_path, delim = "\t", col_names = FALSE)

# Display structure of the files
View(Simulated_FS_Ped27674)
```

# Verify 27674 manipulated .ped File

```{r}
# Define file paths
Manipulated_FS_Ped27674_path <- "/Users/huncho/Desktop/Forensic_Kinship/94_SNP_Simulation/FS_Simulation/FS_Linked_Unrelated/merlin-00003898-27674-replicate_UN.ped"

# Load the Linked file
Manipulated_FS_Ped27674 <- read_delim(Manipulated_FS_Ped27674_path, delim = "\t", col_names = FALSE)

# Display structure of the files
View(Manipulated_FS_Ped27674)
```

# Run MERLIN to extract lnLikelihoods for unrelated cases
# Note: We are not re-simulating — genotypes must remain consistent between FS_Linked and FS_Linked_Unrelated.
# This script loops over all 50,000 manipulated .ped files and saves the lnLikelihoods to individual files.
# All extracted lnLikelihood values are also compiled into a single summary file for downstream processing.

```{bash}
$ nano FS_Linked_Unrelated.sh

#!/bin/bash

# Set working directories
BASE_DIR="/Users/huncho/Desktop/Forensic_Kinship/94_SNP_Simulation/FS_Simulation"
PED_DIR="$BASE_DIR/FS_Linked_Unrelated"
OUTPUT_DIR="$BASE_DIR/FS_Linked_Unrelated"
INPUT_FILES="$BASE_DIR/Input_Files"

# Define marker and frequency files
# Define random seed
MARKER_DAT="$INPUT_FILES/Markers_ForenSeq_SNP.dat"
MARKER_MAP="$INPUT_FILES/Linked_ForenSeqSNP.map"
FREQ_FILE="$INPUT_FILES/WBFreq_94_SNPs_FamLink.freq"
SEED=3898

# Prepare summary file header
OUTFILE="$OUTPUT_DIR/FS_Linked_Unrelated_lnlikelihoods.txt"
echo -e "replicate\tlnlikelihood" > "$OUTFILE"

# Create folder for storing individual lnLikelihood output files
LNLIKELIHOOD_DIR="$OUTPUT_DIR/FS_Linked_Unrelated_lnlikelihoods"
mkdir -p "$LNLIKELIHOOD_DIR"

# Loop through all 50,000 replicates (00001–50000)
for i in $(seq -f "%05g" 1 50000); do
  PED_FILE="$PED_DIR/merlin-00003898-${i}-replicate_UN.ped"
  LNLIKELIHOOD_FILE="$LNLIKELIHOOD_DIR/merlin-00003898-${i}.lnlikelihood"

# Check if .ped file exists before running MERLIN

  if [[ -f "$PED_FILE" ]]; then
    echo "Running Merlin for replicate $i …"

    # Run Merlin to calculate lnlikelihood
    merlin \
      -p "$PED_FILE" \
      -d "$MARKER_DAT" \
      -m "$MARKER_MAP" \
      -f "$FREQ_FILE" \
      -r "$SEED" \
      --likelihood \
      --markerNames \
    > "$LNLIKELIHOOD_FILE" 2>&1

    # Extract the last line containing "likelihood =" from MERLIN output
    LINE=$(grep -i "likelihood.*=" "$LNLIKELIHOOD_FILE" | tail -n1)

    # Extract the numeric value after the "=" sign
    LN=$(echo "$LINE" | cut -d '=' -f 2 | xargs)

    # Append replicate ID and lnLikelihood value to the summary file
    echo -e "${i}\t${LN}" >> "$OUTFILE"

    # If .ped file does not exist, log a warning and skip
  else
    echo "WARNING: $PED_FILE not found, skipping."
  fi
done

# Final message upon completion
echo "All done!  See summary in $OUTFILE and full outputs in $LNLIKELIHOOD_DIR/"
```

```{bash}
# Make the script executable and run it
chmod +x FS_Linked_Unrelated.sh
./FS_Linked_Unrelated.sh
```
