---
title: "FC_Linked"
output: html_document
date: "2025-06-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# NOTE: All simulations, file manipulations, and batch computations (e.g., LR generation, pedigree edits, and data summarisation) were conducted on a HPC system due to local memory and storage constraints.
# Output and intermediate files were transferred to a local environment for visualisation and downstream analysis in R, leveraging RStudio’s improved interactivity and data inspection capabilities.

# Load Packages

```{r}
library(dplyr)
library(readr)
```

# Verify original .ped File

```{r}
# Define file paths
FC_Linked_ped1_path <- "/Users/huncho/Desktop/Forensic_Kinship_Project/10,000_SNPs/FC_Simulation/Input_Files/FC_10000.ped"

# Load the Linked file
Original_FC_ped <- read_delim(FC_Linked_ped1_path, delim = "\t", col_names = FALSE)

# Display structure of the files
View(Original_FC_ped)
```

# Simulation:
# MERLIN was installed in a dedicated conda environment to ensure reproducibility and compatibility with simulation tools.
# The SLURM batch script (merlin_FC_Linked.sh) is configured to simulate 50,000 replicates 
# Outputs are saved with a defined prefix for downstream kinship analysis.

```{bash}
# HPC

# Create a conda environment and install MERLIN (v1.1.2) for reproducible execution
$ conda create -n merlin_env -c bioconda -c 
$ conda-forge merlin=1.1.2 -y
```

```{bash}
# HPC
# Prepare and submit SLURM batch script for MERLIN simulation

$ nano merlin_FC_Linked.sh

#!/bin/bash

# SLURM job settings
#SBATCH --job-name=merlin_FC_Linked
#SBATCH --output=merlin_FC_Linked.out
#SBATCH --error=merlin_FC_Linked.err
#SBATCH --ntasks=1
#SBATCH --cpus-per-task=4
#SBATCH --mem=16G
#SBATCH --time=48:00:00

# Load conda environment
eval "$(conda shell.bash hook)"
conda activate merlin_env

# Move to working directory containing simulation inputs
cd /scratch/users/k21040223/FC_Simulation/FC_Linked

# Simulate MERLIN
merlin -p ../Original_Files/FC_10000.ped -d ../Original_Files/Updated_Kintelligence.dat -m ../Original_Files/Updated_Kintelligence_Map_noX.map -f ../Original_Files/Kintelligence_British_10000.freq -r 3898 --simulate --reruns 50000 --likelihood --information --markerNames --quiet --save --prefix FC_Linked/FC_Linked > likelihoods_FC_Linked_summary.txt

# Submit job to HPC
$ sbatch merlin_FC_Linked.sh
```

# MERLIN outputs per-iteration lnlikelihoods for each chromosome (22 in total)
# So each of the 50,000 simulations generates 22 ln(L) values, one per chromosome
# Need to sum every set of 22 consecutive values (representing one simulation)
# Output the replicate number and total ln(L) to a clean summary file

```{bash}
# HPC

# Create and edit the SLURM batch script
$ nano sum_lnlikelihoods.sh

#!/bin/bash
#SBATCH --job-name=sum_lnlikelihoods
#SBATCH --output=sum_lnlikelihoods.out
#SBATCH --error=sum_lnlikelihoods.err
#SBATCH --ntasks=1
#SBATCH --cpus-per-task=1
#SBATCH --mem=16GB
#SBATCH --time=48:00:00

# Set input/output paths
INFILE="/scratch/users/k21040223/FC_Simulation/FC_Linked/likelihoods_FC_Linked_summary.txt"
OUTFILE="/scratch/users/k21040223/FC_Simulation/FC_Linked/FC_Linked_lnlikelihoods.txt"

# Write header line to output
echo -e "replicate\ttotal_lnlikelihood" > "$OUTFILE"

# AWK script: extract and sum lnL per simulation (22 values per replicate)
awk '
  $0 ~ /likelihood/ {
    # Split line by "=" and take the second part (after the equals sign)
    split($0, parts, "=")
    lnL = parts[2] + 0     # force to numeric (automatically strips spaces, text)
    sum += lnL
    count++

    if (count == 22) {
      printf "%05d\t%.3f\n", ++rep, sum >> "'"$OUTFILE"'"
      sum = 0
      count = 0
    }
  }
' "$INFILE"

echo "Finished. Output written to $OUTFILE"

# Submit the job to SLURM
$ sbatch sum_lnlikelihoods.sh
```

# Verify 1st simulated .ped File

```{r}
# Define file paths
FC_Linked_ped1_path <- "/Users/huncho/Desktop/Forensic_Kinship_Project/10,000_SNPs/FC_Simulation/FC_Linked/merlin-00003898-00001-replicate.ped"

# Load the Linked file
FC_Linked_ped1 <- read_delim(FC_Linked_ped1_path, delim = "\t", col_names = FALSE)

# Display structure of the files
View(FC_Linked_ped1)
```