---
title: "LR"
output: html_document
date: "2025-05-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Packages

```{r}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
```

# Load ln-likelihood summary files for each simulation scenario

```{r}
# Read files

fc_linked <- read.table("/Users/huncho/Desktop/Forensic_Kinship/10,000_SNP_Simulation/FC_Simulation/FC_Linked/FC_Linked_lnlikelihoods.txt", header = TRUE)
fc_linked_unrelated <- read.table("/Users/huncho/Desktop/Forensic_Kinship/10,000_SNP_Simulation/FC_Simulation/FC_Linked_Unrelated/FC_Linked_Unrelated_lnlikelihoods.txt", header = TRUE)

un_linked <- read.table("/Users/huncho/Desktop/Forensic_Kinship/10,000_SNP_Simulation/FC_Simulation/UN_Linked/UN_Linked_lnlikelihoods.txt", header = TRUE)
un_linked_unrelated <- read.table("/Users/huncho/Desktop/Forensic_Kinship/10,000_SNP_Simulation/FC_Simulation/UN_Linked_Unrelated/UN_Linked_Unrelated_lnlikelihoods.txt", header = TRUE)

fc_unlinked <- read.table("/Users/huncho/Desktop/Forensic_Kinship/10,000_SNP_Simulation/FC_Simulation/FC_Unlinked/FC_Unlinked_lnlikelihoods.txt", header = TRUE)
fc_unlinked_unrelated <- read.table("/Users/huncho/Desktop/Forensic_Kinship/10,000_SNP_Simulation/FC_Simulation/FC_Unlinked_Unrelated/FC_Unlinked_Unrelated_lnlikelihoods.txt", header = TRUE)

un_unlinked <- read.table("/Users/huncho/Desktop/Forensic_Kinship/10,000_SNP_Simulation/FC_Simulation/UN_Unlinked/UN_Unlinked_lnlikelihoods.txt", header = TRUE)
un_unlinked_unrelated <- read.table("/Users/huncho/Desktop/Forensic_Kinship/10,000_SNP_Simulation/FC_Simulation/UN_Unlinked_Unrelated/UN_Unlinked_Unrelated_lnlikelihoods.txt", header = TRUE)
```

# Join ln-likelihood tables by replicate
# Each table contains ln-likelihoods for a simulation scenario,
# Merge them into one data frame using the shared 'replicate' column,
# renaming the total_lnlikelihood column to reflect each scenario,

```{r}
all_lnL <- fc_linked %>%
  rename(fc_linked = total_lnlikelihood) %>%
  inner_join(fc_linked_unrelated %>% rename(fc_linked_unr = total_lnlikelihood), by = "replicate") %>%
  inner_join(un_linked %>% rename(un_linked = total_lnlikelihood), by = "replicate") %>%
  inner_join(un_linked_unrelated %>% rename(un_linked_unr = total_lnlikelihood), by = "replicate") %>%
  inner_join(fc_unlinked %>% rename(fc_unlinked = total_lnlikelihood), by = "replicate") %>%
  inner_join(fc_unlinked_unrelated %>% rename(fc_unlinked_unr = total_lnlikelihood), by = "replicate") %>%
  inner_join(un_unlinked %>% rename(un_unlinked = total_lnlikelihood), by = "replicate") %>%
  inner_join(un_unlinked_unrelated %>% rename(un_unlinked_unr = total_lnlikelihood), by = "replicate")
```


# Likelihood Ratio (LR) Calculations
# For each replicate, compute:
# - logLR: difference in ln-likelihoods (log-likelihood ratio)
# - LR: exponentiated value (LR = exp(logLR))
# - log10LR: base-10 transformation for reporting

```{r}
all_lnL <- all_lnL %>%
  mutate(
    # Compute log-likelihood ratios (H1 - H2)
    logLR_FC_linked = fc_linked - fc_linked_unr,
    logLR_UN_linked = un_linked - un_linked_unr,
    logLR_FC_unlinked = fc_unlinked - fc_unlinked_unr,
    logLR_UN_unlinked = un_unlinked - un_unlinked_unr,
    
    # Convert to likelihood ratios (natural scale)
    LR_FC_linked = exp(logLR_FC_linked),
    LR_UN_linked = exp(logLR_UN_linked),
    LR_FC_unlinked = exp(logLR_FC_unlinked),
    LR_UN_unlinked = exp(logLR_UN_unlinked),
    
    # Convert to base-10 scale
    log10LR_FC_linked = logLR_FC_linked / log(10),
    log10LR_UN_linked = logLR_UN_linked / log(10),
    log10LR_FC_unlinked = logLR_FC_unlinked / log(10),
    log10LR_UN_unlinked = logLR_UN_unlinked / log(10)
  )
```

# Export full LR and logLR dataset

```{r}
write.table(all_lnL, "FCvUN_LR_full_output.txt", sep = "\t", row.names = FALSE, quote = FALSE)
```

# Generate summary statistics (min, median, max) for logLR values
# Applies to each of the four comparison conditions

```{r}
summary_stats <- data.frame(
  condition = c("FC_linked", "UN_linked", "FC_unlinked", "UN_unlinked"),
  min = c(
    min(all_lnL$logLR_FC_linked),
    min(all_lnL$logLR_UN_linked),
    min(all_lnL$logLR_FC_unlinked),
    min(all_lnL$logLR_UN_unlinked)
  ),
  median = c(
    median(all_lnL$logLR_FC_linked),
    median(all_lnL$logLR_UN_linked),
    median(all_lnL$logLR_FC_unlinked),
    median(all_lnL$logLR_UN_unlinked)
  ),
  max = c(
    max(all_lnL$logLR_FC_linked),
    max(all_lnL$logLR_UN_linked),
    max(all_lnL$logLR_FC_unlinked),
    max(all_lnL$logLR_UN_unlinked)
  )
)

# Export summary statistics to file
write.table(summary_stats, "FCvUN_logLR_summary_stats.txt", sep = "\t", row.names = FALSE, quote = FALSE)
```

# PLOTTING: Density Plot of log10LRs for FC vs UN (9991 SNPs)

```{r}
# Prepare plotting data: extract log10LRs and reshape into long format
plot_data <- all_lnL %>%
  transmute(
    FC_linked    = log10LR_FC_linked,
    FC_unlinked  = log10LR_FC_unlinked,
    UN_linked    = log10LR_UN_linked,
    UN_unlinked  = log10LR_UN_unlinked
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Group",
    values_to = "log10LR"
  )

# Create density plot for all log10LRs
p <- ggplot(plot_data, aes(x = log10LR, colour = Group, fill = Group)) +
  geom_density(alpha = 0.5, size = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "black") +
  labs(
    title = "Distribution of log10LRs for FC vs UN (9991 SNPs)",
    x = "log10LR",
    y = "Density"
  ) +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    legend.position = "right"
  )

# Display plot
print(p)

# Save as PDF
ggsave(
  filename = "FC_vs_UN_9991SNPs_log10LR.pdf",
  plot = p,
  device = "pdf",
  width = 8,
  height = 6,
  dpi = 300
)
```

# CAPPED VERSION: Density Plot of log10LRs with Y-Axis Restricted

```{r}
# Prepare plotting data: extract log10LRs and reshape into long format
plot_data_fc <- all_lnL %>%
  transmute(
    FC_linked    = log10LR_FC_linked,
    FC_unlinked  = log10LR_FC_unlinked,
    UN_linked    = log10LR_UN_linked,
    UN_unlinked  = log10LR_UN_unlinked
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Group",
    values_to = "log10LR"
  ) %>%
  

# Create capped plot with y-axis limited
p_fc <- ggplot(plot_data_fc, aes(x = log10LR, colour = Group, fill = Group)) +
  geom_density(alpha = 0.5, size = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "black") +
  coord_cartesian(ylim = c(0, 0.05)) +
  labs(
    title = "Distribution of Capped log10LRs for FC vs UN (9991 SNPs)",
    x = "log10LR",
    y = "Density"
  ) +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    legend.position = "right"
  )

# Display capped plot
print(p_fc)

# Save capped plot
ggsave(
  filename = "FC_vs_UN_9991SNPs_log10LR_ycapped.pdf",
  plot = p_fc,
  width = 8,
  height = 6,
  dpi = 300
)
```

