---
title: "FC_Unlinked"
output: html_document
date: "2025-06-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Packages

```{r}
library(dplyr)
library(readr)
```

# Run MERLIN to calculate lnLikelihoods assuming unlinked markers
# This script uses the unlinked map file to ignore genetic linkage when computing likelihoods
# Output lnLikelihoods are generated for each replicate in the FC_Linked .ped dataset

```{bash}
#HPC 

$ nano FC_Unlinked.sh

#!/bin/bash
#SBATCH --job-name=merlin_FC_Unlinked
#SBATCH --output=merlin_FC_Unlinked.out
#SBATCH --error=merlin_FC_Unlinked.err
#SBATCH --ntasks=1
#SBATCH --cpus-per-task=4
#SBATCH --mem=16G
#SBATCH --time=48:00:00

# Load conda environment
eval "$(conda shell.bash hook)"
conda activate merlin_env

# Set working directories
BASE_DIR="/scratch/users/k21040223/FC_Simulation"
PED_DIR="$BASE_DIR/FC_Linked"
OUTPUT_DIR="$BASE_DIR/FC_Unlinked"
ORIGINAL_FILES="$BASE_DIR/Original_Files"

# Define marker and frequency files
# Define random seed
MARKER_DAT="$ORIGINAL_FILES/Updated_Kintelligence.dat"
MARKER_MAP="$ORIGINAL_FILES/Updated_Kintelligence_Map_noX.map"
FREQ_FILE="$ORIGINAL_FILES/Kintelligence_British_10000.freq"
SEED=3898

# Prepare summary file header
OUTFILE="$OUTPUT_DIR/FC_Unlinked_lnlikelihoods.txt"
echo -e "replicate\ttotal_lnlikelihood" > "$OUTFILE"

# Create folder for storing individual lnLikelihood output files
LNLIKELIHOOD_DIR="$OUTPUT_DIR/FC_Unlinked_lnlikelihoods"
mkdir -p "$LNLIKELIHOOD_DIR"

# Loop through all 50,000 replicates (00001–50000)
for i in $(seq -f "%05g" 1 50000); do
  PED_FILE="$PED_DIR/merlin-00003898-${i}-replicate.ped"
  LNLIKELIHOOD_FILE="$LNLIKELIHOOD_DIR/merlin-00003898-${i}.lnlikelihood"

# Check if .ped file exists before running MERLIN

  if [[ -f "$PED_FILE" ]]; then
    echo "Running Merlin for replicate $i …"

    # Run Merlin to calculate lnlikelihood
    merlin \
      -p "$PED_FILE" \
      -d "$MARKER_DAT" \
      -m "$MARKER_MAP" \
      -f "$FREQ_FILE" \
      -r "$SEED" \
      --likelihood \
      --markerNames \
      > "$LNLIKELIHOOD_FILE" 2>&1

    # Extract all lnlikelihood values (one per chromosome), sum them
      LN_SUM=$(grep -i "lnlikelihood for" "$LNLIKELIHOOD_FILE" \
      | awk -F'= *' '
          {
            split($2, tokens, " ")
            sum += tokens[1]
          }
          END {
            printf "%.3f", sum
          }
        ')

    # Append to the summary file
    echo -e "${i}\t${LN_SUM}" >> "$OUTFILE"

  # If .ped file does not exist, log a warning and skip
  else
    echo "WARNING: $PED_FILE not found, skipping."
  fi
done

# Final message upon completion
echo "All done! See summary in $OUTFILE and full outputs in $LNLIKELIHOOD_DIR/"

$ sbatch FC_Unlinked.sh
```

# NOTE:
# The FC_Linked_Unrelated and FC_Unlinked_Unrelated conditions both represent two unrelated individuals with no pedigree structure.

# Since the underlying genotypes and likelihood model assumptions are identical,the ln-likelihood values computed by MERLIN are mathematically guaranteed to be the same.

# To save computational time, ln-likelihoods were only calculated once (for FC_Linked_Unrelated) and reused for FC_Unlinked_Unrelated where applicable.
