---
title: "UN_Linked_Unrelated"
output: html_document
date: "2025-06-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Packages 

```{r}
library(dplyr)
library(readr)
```

# The 50,000 .ped files from the UN_Linked simulation are manipulated to simulate unrelated individuals.
# This is done by:
# 1) Breaking familial relationships by setting columns 3 (Mother ID) and 4 (Father ID) to 0.
# 2) Removing extra rows/columns added during simulation.
# 3) Standardising each .ped file to contain only two unrelated individuals:
#    - First 5 columns fixed
#    - Remaining columns: genotypes from each simulated .ped file

```{bash}
# HPC

$ nano UN_Peds_Unrelated.R
```

```{r}
library(dplyr)
library(readr)

# Set the directories, input directory is where the UN_Linked simulated .peds are stored and the output directory is where the manipulated .peds will be stored
in_dir  <- "/scratch/users/k21040223/FS_Simulation/UN_Linked"
out_dir <- "/scratch/users/k21040223/FS_Simulation/UN_Linked_Unrelated"

# Create the output directory if it doesn’t exist
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

# Parse command line arguments
args <- commandArgs(trailingOnly = TRUE)
start_index <- as.integer(args[1])  # e.g., 1
end_index   <- as.integer(args[2])  # e.g., 500

# Define the transformation as a function
process_ped <- function(in_ped, out_ped) {
  df <- read_delim(in_ped, delim = "\t", col_names = FALSE, show_col_types = FALSE)
  colnames(df) <- sprintf("X%d", seq_len(ncol(df)))

  df_mod <- df %>%
    # Set X3 and X4 (mother and father IDs) to 0 to break any pedigree structure
    mutate(X3 = 0, X4 = 0) %>%

    # Shift genotypes from rows 3 and 4 to become the first two rows (representing two unrelated individuals)
    {
      tmp  <- .
      cols <- 6:ncol(tmp)
      tmp[1, cols] <- tmp[3, cols]
      tmp[2, cols] <- tmp[4, cols]
      tmp
    } %>%

    # Drop all original rows except the newly reassigned unrelated pair (keep only rows 1 and 2)
    slice(-c(3, 4))

  # Write the modified .ped file to the output directory
  write_delim(df_mod, out_ped, delim = "\t", col_names = FALSE)
}

# Loop through specified replicates and apply transformation
for (i in start_index:end_index) {
  id       <- sprintf("%05d", i)
  in_file  <- file.path(in_dir,  sprintf("UN_ped_%s-fixed.ped", id))
  out_file <- file.path(out_dir, sprintf("UN_ped_%s-Unrelated.ped", id))

  if (!file.exists(in_file)) {
    warning(sprintf("Missing: %s", in_file))
    next
  }

  process_ped(in_file, out_file)
}
```

```{bash}
# HPC

# Create and edit the SLURM batch script
$ nano UN_Peds_Unrelated_Array.sh

#!/bin/bash
#SBATCH --job-name=UN_Peds_Unrelated_Array
#SBATCH --array=0-99
#SBATCH --output=logs/UN_Peds_Unrelated_Array_%A_%a.out
#SBATCH --error=logs/UN_Peds_Unrelated_Array_%A_%a.err
#SBATCH --time=48:00:00
#SBATCH --mem=16G

module load r

# Each task processes 500 files
START=$(( SLURM_ARRAY_TASK_ID * 500 + 1 ))
END=$(( START + 499 ))

Rscript UN_Peds_Unrelated.R $START $END

$ sbatch UN_Peds_Unrelated_Array.sh
```

# Verify structure of 1st modified .ped

```{r}
# Define file paths
Simulated_UN_ped1_modified_UN_path <- "/Users/huncho/Desktop/Forensic_Kinship/10,000_SNP_Simulation/FS_Simulation/UN_Linked_Unrelated/UN_ped_00001-Unrelated.ped"

# Load the Linked file
Simulated_UN_ped1_modified_UN <- read_delim(Simulated_UN_ped1_modified_UN_path, delim = "\t", col_names = FALSE)

# Display structure of the files
View(Simulated_UN_ped1_modified_UN)
```

# Run MERLIN to extract lnLikelihoods for unrelated cases (H2)
# Note: We are not re-simulating — genotypes must remain consistent between UN_Linked and UN_Linked_Unrelated.
# This script loops over all 50,000 manipulated .ped files and saves the lnLikelihoods to individual files.
# All extracted lnLikelihood values are also compiled into a single summary file for downstream processing.

```{bash}
# HPC

$ nano UN_Linked_Unrelated.sh

#!/bin/bash
#SBATCH --job-name=merlin_UN_Linked_Unrelated
#SBATCH --output=merlin_UN_Linked_Unrelated.out
#SBATCH --error=merlin_UN_Linked_Unrelated.err
#SBATCH --ntasks=1
#SBATCH --cpus-per-task=4
#SBATCH --mem=16G
#SBATCH --time=48:00:00

# Set working directories
BASE_DIR="/scratch/users/k21040223/FS_Simulation"
PED_DIR="$BASE_DIR/UN_Linked_Unrelated"
OUTPUT_DIR="$BASE_DIR/UN_Linked_Unrelated"
ORIGINAL_FILES="$BASE_DIR/Original_Files"

# Define marker and frequency files
# Define random seed
MARKER_DAT="$ORIGINAL_FILES/Updated_Kintelligence.dat"
MARKER_MAP="$ORIGINAL_FILES/Updated_Kintelligence_Map_noX.map"
FREQ_FILE="$ORIGINAL_FILES/Kintelligence_British_10000.freq"
SEED=3898

# Prepare summary file header
OUTFILE="$OUTPUT_DIR/UN_Linked_Unrelated_lnlikelihoods.txt"
echo -e "replicate\ttotal_lnlikelihood" > "$OUTFILE"

# Create folder for storing individual lnLikelihood output files
LNLIKELIHOOD_DIR="$OUTPUT_DIR/lnlikelihoods"
mkdir -p "$LNLIKELIHOOD_DIR"

# Loop through all 50,000 replicates (00001–50000)
for i in $(seq -f "%05g" 1 50000); do
  PED_FILE="$PED_DIR/UN_ped_${i}-Unrelated.ped"
  LNLIKELIHOOD_FILE="$LNLIKELIHOOD_DIR/merlin-00003898-${i}.lnlikelihood"

# Check if .ped file exists before running MERLIN

  if [[ -f "$PED_FILE" ]]; then
    echo "Running Merlin for replicate $i …"

     # Run Merlin to calculate lnlikelihood
    merlin \
      -p "$PED_FILE" \
      -d "$MARKER_DAT" \
      -m "$MARKER_MAP" \
      -f "$FREQ_FILE" \
      -r "$SEED" \
      --likelihood \
      --markerNames \
      > "$LNLIKELIHOOD_FILE" 2>&1
      
      # Extract all lnlikelihood values (one per chromosome), sum them
      LN_SUM=$(grep -i "lnlikelihood for" "$LNLIKELIHOOD_FILE" \
      | awk -F'= *' '
          {
            split($2, tokens, " ")
            sum += tokens[1]
          }
          END {
            printf "%.3f", sum
          }
        ')

    # Append to the summary file
    echo -e "${i}\t${LN_SUM}" >> "$OUTFILE"

  # If .ped file does not exist, log a warning and skip
  else
    echo "WARNING: $PED_FILE not found, skipping."
  fi
done

# Final message upon completion
echo "All done! See summary in $OUTFILE and full outputs in $LNLIKELIHOOD_DIR/"

$ sbatch UN_Linked_Unrelated.sh
```

